Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2021-08-24 summary.tex
fixed a couple typos

\begin{abstract}
We connect two random graph models, the Popularity Adjusted Block Model
(PABM) and the Generalized Random Dot Product Graph (GRDPG),
by demonstrating that a PABM is a GRDPG in which communities correspond to 
mutually orthogonal subspaces of latent vectors. 
This insight allows us to construct new algorithms for community detection
and parameter estimation with the PABM, 
as well as improve an existing algorithm 
that relies on Sparse Subspace Clustering. 
Using established asymptotic properties of 
Adjacency Spectral Embedding for the GRDPG, 
we derive asymptotic properties of these algorithms. 
In particular, we demonstrate that the absolute number of 
community detection errors tends to zero as 
the number of graph vertices tends to infinity. 
Simulation experiments illustrate these properties. 
\end{abstract}

@@

Statistical inference on random graphs requires a suitable probability model.  A general probability model for unweighted and undirected graphs is the Bernoulli Graph (also known as the inhomogeneous Er\"{o}s-R\'{e}nyi model), which assumes that edges occur as independent Bernoulli trials.  A Bernoulli Graph is characterized by an edge probability matrix $P=[P_{ij}]$, where an edge between vertices $i$ and $j$ occurs with success probability $P_{ij}$.  A trivial example of a Bernoulli Graph is the Erd\"{o}s-R\'{e}nyi model proposed by \citet{Gilbert:1959}, in which the vertices of the random graph are fixed and possible edges occur independently with fixed probability $P_{ij}=p$.  The requirement that $p$ is fixed is too strong for most applications, and various researchers have weakened that requirement in various ways.  The present work relates two lines of generalization.

@@

Another type of Bernoulli Graph was proposed by \citet*{10.1007/978-3-540-77004-6_11}.  A Random Dot Product Graph (RDPG) specifies that each vertex corresponds to a latent position vector in Euclidean space and that the probability of an edge between two vertices is the dot product of their latent position vectors.  Thus, if the latent positions $x_1,\ldots,x_n \in \mathbb{R}^d$ and $X = [\begin{array}{ccc} x_1 & \cdots & x_n \end{array} ]^\top$, then the edge probability matrix is $P = XX^\top$.  Clearly, any Bernoulli Graph with positive semidefinite $P$ is an RDPG.  The positive definite Euclidean inner product in the RDPG model was replaced by an indefinite inner product in \citet{rubindelanchy2017consistency}, resulting in the {\em Generalized}\/ RDPG (GRDPG).

@@

\begin{comment}

Statistical inference on random graphs involves the
partitioning of a graph into disconnected subgraphs or clusters. This is
often motivated by the assumption that there exist underlying and
unobserved communities to which each vertex of the graph belongs, and
edges between pairs of vertices are determined by drawing from a
probability distribution based on the community relationships between
each pair. The goal of this analysis then is population community
detection, or the recovery of the true underlying community labels for
each vertex, up to permutation, with some additional parameter
estimation being of possible interest, assuming some underlying
probability model.

A general family of models that describe unweighted and undirected graphs is 
the Bernoulli Graph, which is characterized by edge probability matrix $P$. 
Edges between each pair of vertices $i$ and $j$ are drawn 
as a Bernoulli trial with probability $P_{ij}$, for each 
$0 \leq i < j \leq n$ and $n$ is the number of vertices in the graph. 
Community detection methods often restrict this type of model 
by assigning each vertex a community label $z_1, ..., z_n$ 
and then setting each $P_{ij}$ as a function of the labels $z_i$ and $z_j$. 
Such models are called Block Models.

One popular type of Block Model is the Stochastic Block Model (SBM),
first proposed by \citet{doi:10.1080/0022250X.1971.9989788}, 
which restricts each edge probability $P_{ij}$ as only depending on the labels 
$z_i$ and $z_j$, i.e., $P_{ij} = \omega_{z_i, z_j}$. 
Various generalizations of the SBM have since been utilized, 
including the Degree-Corrected Block Model (DCBM), 
introduced by \citet{Karrer_2011}, 
which assigns an additional parameter $\theta_i$ to each vertex 
and sets $P_{ij} = \theta_i \theta_j \omega_{z_i, z_j}$. 
The Popularity Adjusted Block Model (PABM) was then introduced by
\citet*{307cbeb9b1be48299388437423d94bf1} as a generalization of the
DCBM to address the heterogeneity of edge probabilities within and
between communities while still maintaining distinct community
structure.

Parallel to the families of Block Models is another type of Bernoulli Graph 
called the Random Dot Product Graph (RDPG), proposed by
\citet*{10.1007/978-3-540-77004-6_11}. 
Under this model, each vertex of the graph
can be represented by a vector in some latent space such that the edge
probability between any pair of vertices is given by their corresponding
dot product in the latent space, i.e., given a latent positions
\(x_1, ..., x_n \in \mathbb{R}^d\), the edge probability matrix is
\(P = X X^\top\) where 
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\). 
It can be shown that all Bernoulli Graphs with 
positive semidefinite edge probability matrix $P$ is a type of RDPG. 
For instance, the assortative
SBM is equivalent to a special case of the RDPG for which all vertices
of a given community share the same position in the latent space
\cite{lyzinski2014}. A similar property exists for the DCBM, 
which is equivalent to a RDPG in which vertices that belong to the same 
community all lie in a line segment in the latent space 
\cite{lyzinski2014, rubindelanchy2017consistency}. 
Because the edge probability matrix for the PABM is not positive semidefinite, 
we instead show that it is a special case of the 
\emph{Generalized} Random Dot Product Graph (GRDPG) 
in order to analyze the PABM as a latent space model 
for community detection and parameter estimation.

\end{comment}

@@

for a set of $K^2$ popularity vectors $\{\lambda^{(st)}\}_{1 \leq
  s \leq K, 1 \leq t \leq K}$ where each
$\lambda^{(st)}$ is a column vector of dimension $n_s$. Then a graph $G$
is drawn from a PABM with parameters
$\{\lambda^{(st)}\}$ if its adjacency matrix $A$
satisfies $A \sim \mathrm{Bernoulli}(P)$.
\end{definition}

@@

Our central result states that the Popularity Adjusted Block Model 
is a special case of the Generalized Random Dot Product Graph. 
In particular, the PABM with $K$ communities is a GRDPG 
for which the communities are represented by 
mutually orthogonal $K$-dimensional subspaces of the latent space. 
This result extends previous results that connected the Stochastic Block Model and the Degree Corrected Block Model to Random Dot Product Graphs. 
Replacing RDPGs with GRDPGs is a critical step in this line of research, as a PABM is not necessarily a RDPG.

Because all Bernoulli Graphs are GRDPGs, it should be possible to invent and study new families of Bernoulli Graphs by characterizing them as special cases of GRDPGs and expliting the latent structures that define them. 
The present work illustrates the powr of this approach. 
We recover the latent structure of the PABM by Adjacency Spectral Embedding, then exploit that structure to improve statistical inference. 
Exploiting the fact that PABM communities correspond to orthogonal subspaces, we propose Orthogonal Spectral Clustering for community detection and demonstrate that the number of misclassified vertices approaches zero with high probability as the size of the graph increases. 
This is a stronger result than previously proposed algorithms \citep{307cbeb9b1be48299388437423d94bf1}, which only guarantee that the error rate (and not count) approaches zero asymptotically. 
Parameter estimation can be performed in a similar fashion using the ASE, for which we also prove that the per-parameter error approaches zero asymptotically.

>>>>

\begin{abstract}
We connect two random graph models, the Popularity Adjusted Block Model
(PABM) and the Generalized Random Dot Product Graph (GRDPG),
by demonstrating that a PABM is a GRDPG in which communities correspond to 
mutually orthogonal subspaces of latent vectors. 
This insight allows us to construct new algorithms for community detection
and parameter estimation for the PABM, 
as well as improve an existing algorithm 
that relies on Sparse Subspace Clustering. 
Using established asymptotic properties of 
Adjacency Spectral Embedding for the GRDPG, 
we derive asymptotic properties of these algorithms. 
In particular, we demonstrate that the absolute number of 
community detection errors tends to zero as 
the number of graph vertices tends to infinity. 
Simulation experiments illustrate these properties. 
\end{abstract}

@@

Statistical inference on random graphs requires a suitable probability model. 
A general probability model for unweighted and undirected graphs is the Bernoulli Graph (also known as the inhomogeneous Erd\"{o}s-R\'{e}nyi model), which assumes that edges occur as independent Bernoulli trials.  A Bernoulli Graph is characterized by an edge probability matrix $P=[P_{ij}]$, where an edge between vertices $i$ and $j$ occurs with success probability $P_{ij}$.  A trivial example of a Bernoulli Graph is the (homogeneous) Erd\"{o}s-R\'{e}nyi model proposed by \citet{Gilbert:1959}, in which the vertices of the random graph are fixed and possible edges occur independently with fixed probability $P_{ij}=p$.  The requirement that $p$ is fixed is too strong for most applications, and various researchers have weakened that requirement in various ways.  The present work relates two lines of generalization.

@@

Another type of Bernoulli Graph was proposed by \citet*{10.1007/978-3-540-77004-6_11}.  A Random Dot Product Graph (RDPG) specifies that each vertex corresponds to a latent position vector in Euclidean space and that the probability of an edge between two vertices is the dot product of their latent position vectors.  Thus, if the latent positions are $x_1,\ldots,x_n \in \mathbb{R}^d$ and $X = [\begin{array}{ccc} x_1 & \cdots & x_n \end{array} ]^\top$, then the edge probability matrix is $P = XX^\top$.  Clearly, any Bernoulli Graph with positive semidefinite $P$ is an RDPG.  The positive definite Euclidean inner product in the RDPG model was replaced by an indefinite inner product in \citet{rubindelanchy2017consistency}, resulting in the {\em Generalized}\/ RDPG (GRDPG).

@@

@@

for a set of $K^2$ popularity vectors 
$\{\lambda^{(st)}\}_{1 \leq s \leq K, 1 \leq t \leq K}$ where each
$\lambda^{(st)}$ is a column vector of dimension $n_s$. Then a graph $G$
is drawn from a PABM with parameters
$\{\lambda^{(st)}\}$ if its adjacency matrix $A$
satisfies $A \sim \mathrm{BernoulliGraph}(P)$.
\end{definition}

@@

Our central result states that the Popularity Adjusted Block Model 
is a special case of the Generalized Random Dot Product Graph. 
In particular, the PABM with $K$ communities is a GRDPG 
for which the communities are represented by 
mutually orthogonal $K$-dimensional subspaces 
of the $K^2$-dimensional latent space. 
This result extends previous results that connected 
the Stochastic Block Model and the Degree Corrected Block Model 
to Random Dot Product Graphs. 
Replacing RDPGs with GRDPGs is a critical step in this line of research, 
as a PABM is not necessarily a RDPG.

Because all Bernoulli Graphs are GRDPGs, it should be possible to invent and study new families of Bernoulli Graphs by characterizing them as special cases of GRDPGs and expliting the latent structures that define them. 
The present work illustrates the power of this approach. 
We recover the latent structure of the PABM by Adjacency Spectral Embedding, then exploit that structure to improve statistical inference. 
Exploiting the fact that PABM communities correspond to orthogonal subspaces, we propose Orthogonal Spectral Clustering for community detection and demonstrate that the number of misclassified vertices approaches zero with high probability as the size of the graph increases. 
This is a stronger result than previously proposed algorithms \citep{307cbeb9b1be48299388437423d94bf1}, which only guarantee that the error rate (and not count) approaches zero asymptotically. 
Parameter estimation can be performed in a similar fashion using the ASE, for which we also prove that the per-parameter error approaches zero asymptotically.
